---
title: Compound Annotation and Structure Collection Datasets
author: "Authors: Yuzhu Duan, Dan Evans, Jamison McCorrison, Nicholas Schork, Kevin Horan, Austin Leong, Siddharth Sai and Thomas Girke"
date: "Last update: `r format(Sys.time(), '%d %B, %Y')`" 
always_allow_html: yes
output:
  BiocStyle::html_document:
    toc_float: true
    code_folding: show
vignette: >
  %\VignetteIndexEntry{compoundCollectionData}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
fontsize: 14pt
bibliography: bibtex.bib
---

<!-- <style> -->
<!-- pre code { -->
<!--   white-space: pre !important; -->
<!--   overflow-x: scroll !important; -->
<!--   word-break: keep-all !important; -->
<!--   word-wrap: initial !important; -->
<!-- } -->
<!-- </style> -->

<!---
- Compile from command-line
Rscript -e "rmarkdown::render('compoundCollectionData.Rmd', c('BiocStyle::html_document', 'pdf_document')); knitr::knit('compoundCollectionData.Rmd', tangle=TRUE)"
-->

# Overview

This package contains annotation and structure datasets for compounds in 
[DrugAge](https://genomics.senescence.info/drugs/), 
[DrugBank](https://www.drugbank.ca/), [CMAP02](https://portals.broadinstitute.org/cmap/) 
and [LINCS](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE92742) databases.
The actual datasets are stored in `AnnotationHub`. 

# Install and load package

The following shows how to install the package with the `install_github`
function of the `devtools` package.

```{r pkg_install, eval=FALSE}
library(devtools)
install_github("yduan004/compoundCollectionData")
library(compoundCollectionData)
```

# Dataset Description
The description of the 5 datasets in this package is as follows

## SQLite annotation database
It is a SQLite database storing compound annotation tables for DrugAge, DrugBank,
CMAP02 and LINCS, respectively. It also contains an ID mapping table of ChEMBL IDs
to IDs of individual databases.

The SQLite annotation database can be loaded into R as follows:
```{r sql, eval=FALSE}
library(AnnotationHub)
ah <- AnnotationHub()
query(ah, c("compoundCollectionData", "annot"))
annot_path <- ah[[""]]
library(RSQLite)
conn <- dbConnect(SQLite(), annot_path)
dbListTables(conn)
dbDisconnect(conn)
```

## DrugAge SDF
It is an SDF (Structure-Data File) file storing molecular structures of DrugAge
comopunds. The source annotation file of DrugAge is downloaded from
[here](http://genomics.senescence.info/drugs/dataset.zip)
The extracted csv file only contains drug names, without id mappings to external 
resources such as [PubChem](https://pubchem.ncbi.nlm.nih.gov/) or 
[ChEMBL](https://www.ebi.ac.uk/chembl/). The extracted `drugage.csv` file was 
processed by the `processDrugage` function from the 
[drugTargetInteractions](https://github.com/longevity-consortium/LC_Chemoinformatics/tree/master/Rpackages/drugTargetInteractions) 
package. The missing IDs were then added manually. The result DrugAge annotation table
with external ID mappings is available at https://bit.ly/3dCWKWo. This annotation
table along with the id-mapping table (ChEMBL ID to DrugAge internal id) were
stored in the compoundCollection SQLite database. The drug structures were
obtained from PubChem CIDs by `getIds` function from the `ChemmineR` Bioconductor
package, then written to the `drugage.sdf` file.

The DrugAge compound structures can be loaded into R as follows:
```{r da, eval=FALSE}
query(ah, c("compoundCollectionData", "drugage"))
da_path <- ah[[""]]
da_sdfset <- ChemmineR::read.SDFset(da_path)
```

## DrugBank SDF
This SDF file stores structures of compounds in [DrugBank](https://www.drugbank.ca/).
The full DrugBank xml database was downloaded from [here](https://www.drugbank.ca/releases/latest).
The most recent release version at this time is 5.1.5.  
The extracted xml file was procesed by the [drugbankR](https://github.com/yduan004/drugbankR) package.
The result DrugBank annotation table was then stored in the `compoundCollection` 
SQLite database. The DrugBank to ChEMBL id mappings were obtained from 
[UniChem](ftp://ftp.ebi.ac.uk/pub/databases/chembl/UniChem/data/wholeSourceMapping/src_id1/src1src2.txt.gz).
The DrugBank SDF file was downloaded from [here](https://www.drugbank.ca/releases/latest#structures)
and made some validity checks and modifications via utilities in the `ChemmineR` package.

The DrugBank compound structures can be loaded into R as follows:
```{r db, eval=FALSE}
query(ah, c("compoundCollectionData", "drugbank"))
db_path <- ah[[""]]
db_sdfset <- ChemmineR::read.SDFset(db_path)
```

## CMAP SDF
The CMAP compound annotation table and its structures were obtained from the
[longevityDrugs](https://github.com/tgirke/longevityDrugs) package. 
Since this table only contains PubChem CID, the ChEMBL ids were added via
PubChem CID to ChEMBL id mappings from
[UniChem](ftp://ftp.ebi.ac.uk/pub/databases/chembl/UniChem/data/wholeSourceMapping/src_id1/src1src22.txt.gz).
The CMAP internal IDs were made for ChEMBL to CMAP id mappings.

The DrugAge compound structures can be loaded into R as follows:
```{r cmap, eval=FALSE}
query(ah, c("compoundCollectionData", "cmap"))
cmap_path <- ah[[""]]
cmap_sdfset <- ChemmineR::read.SDFset(cmap_path)
```

## LINCS SDF

The LINCS compound annotation table was downloaded from 
[GEO](ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE92nnn/GSE92742/suppl/GSE92742_Broad_LINCS_pert_info.txt.gz)
where only compounds were selected. The LINCS ids were mapped to ChEMBL ids via 
inchi key. The LINCS compounds structures were obtained from PubChem CIDs via 
the `getIDs` function from the `ChemmineR` package.

The DrugAge compound structures can be loaded into R as follows:
```{r lincs, eval=FALSE}
query(ah, c("compoundCollectionData", "lincs"))
lincs_path <- ah[[""]]
lincs_sdfset <- ChemmineR::read.SDFset(lincs_path)
```

